---
title: "DNAtranslate (R Package)"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

> **Note:** This is a practice package built for learning purposes.

## Overview

`DNAtranslate` is a small R package I created to learn the package-development workflow. It provides simple helpers for importing DNA sequences, reversing them, and translating them into protein sequences. It is not intended for production use—well-established packages handle these tasks far more robustly. For importing sequences in particular, functions such as `readFastq` preserve considerably more information.

Here is a [link](https://github.com/ADC-405-S26/DNAtranslate) to the full source code

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("ADC-405-S26/DNAtranslate")
```

## Examples

The examples below shows how to import a DNA sequence, reverse it, and translate it into a protein sequence.

### Importing DNA sequences

```{r readSequences}
library(DNAtranslate)

# Import a .txt file formatted as follows:
#   >Sequence Name
#   ATGCATGCATGCATGCATGCATGC
#   ATGCATGCATGC...
# The sequence name must be preceded by a ">" and followed by a new line.

# Fetch the package's built-in dataset for testing purposes:
path <- system.file("extdata", "exampleSequences.txt", package = "DNAtranslate")
sequencesDF <- readSequences(path)
head(sequencesDF)
```

### Reversing a sequence

```{r reverseSequence}
library(DNAtranslate)

# Flip a specified sequence within the data frame.
reverse <- reverseSequence(exampleSequences, "HisD TA1538 Mutant Genotype")
head(reverse)
# The sequence named "HisD TA1538 Mutant Genotype" is now reversed.
```

### Translating DNA to protein

```{r translateDNA}
library(DNAtranslate)

# Add a new column containing the protein sequence that corresponds to each DNA sequence.
proteinSequence <- translateDNA(exampleSequences, offset = 0)
head(proteinSequence)
```
