---
title: "SQL Project: Analyzing UFO Sightings"
toc: true
---

*Joint work with Joseph Dee. Source code is available on [GitHub](https://github.com/ADC-405-S26/Project_2_SQL_Team_13/tree/main).*

This project builds a small analytical database from the [TidyTuesday UFO sightings dataset](https://github.com/rfordatascience/tidytuesday/blob/main/data/2023/2023-06-20/readme.md) and uses SQL to explore when, where, and how sightings were reported.

```{r setup}
#| message: false
#| warning: false
#| include: false
library(tidyverse)
library(dbplyr)
library(RMariaDB)
library(duckdb) # R package for building a DuckDB database locally.
```

## Accessing the data

The data consists of three CSV files: `ufo_sightings.csv`, `places.csv`, and `day_parts_map.csv`. `ufo_sightings` shares `city` and `state` with `places`, and shares its date fields with `day_parts_map`, which lets the tables be joined together. After loading the files, we remove duplicate rows so that each table has a clean key, then write the cleaned data back out for import into the database.

```{r}
#| message: false
#| warning: false
raw_ufo_sightings <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2023/2023-06-20/ufo_sightings.csv')
raw_places <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2023/2023-06-20/places.csv')
day_parts_map <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2023/2023-06-20/day_parts_map.csv')

glimpse(raw_ufo_sightings) # `shape` has many NAs; `day_part` has some NAs
glimpse(raw_places)        # `alternative_city_names` and `elevation_m` have NAs
glimpse(day_parts_map)     # No NAs

# `places` contains ~625 rows with duplicate latitude/longitude pairs.
places <- raw_places %>% distinct(latitude, longitude, .keep_all = TRUE)

# We use (reported_date_time_utc, city) as the key for sightings.
ufo_sightings <- raw_ufo_sightings %>% distinct(reported_date_time_utc, city, .keep_all = TRUE)

write_csv(ufo_sightings, "ufo_sightings.csv")
write_csv(places, "places.csv")
write_csv(day_parts_map, "day_parts_map.csv")
```

## Designing the database

We store the data in a local DuckDB database and define an explicit schema for each table, including column types, nullability, and primary keys. Column types follow the [DuckDB data type reference](https://duckdb.org/docs/current/sql/data_types/overview).

```{r}
# Use an in-memory DuckDB database. Every table is rebuilt from the CSV files
# on each render, so nothing needs to persist to disk. An in-memory database
# also avoids the file-lock error that occurs on Windows when a previous
# connection to a "UFO_db" file is still open.
con_UFO <- DBI::dbConnect(duckdb::duckdb(), dbdir = ":memory:")
```

```{sql connection = con_UFO}
DROP TABLE IF EXISTS ufo_sightings;
DROP TABLE IF EXISTS places;
DROP TABLE IF EXISTS day_parts_map;
```

The `ufo_sightings` table uses a composite primary key of `reported_date_time_utc` and `city`. Default values are left unset.

```{sql connection = con_UFO}
CREATE TABLE ufo_sightings (
  reported_date_time TIMESTAMP NOT NULL,
  reported_date_time_utc TIMESTAMP NOT NULL,
  posted_date DATE NOT NULL,
  city VARCHAR NOT NULL,
  state VARCHAR,
  country_code VARCHAR NOT NULL,
  shape VARCHAR,
  reported_duration VARCHAR NOT NULL,
  duration_seconds BIGINT NOT NULL,
  summary VARCHAR,
  has_images BOOLEAN NOT NULL,
  day_part VARCHAR,
  PRIMARY KEY (reported_date_time_utc, city)
);
```

```{sql connection=con_UFO}
COPY ufo_sightings FROM 'ufo_sightings.csv' (FORMAT CSV, HEADER, NULL 'NA');
```

The `places` table uses latitude and longitude as its primary key.

```{sql connection = con_UFO}
CREATE TABLE places (
  city VARCHAR NOT NULL,
  alternate_city_names VARCHAR,
  state VARCHAR,
  country VARCHAR NOT NULL,
  country_code VARCHAR NOT NULL,
  latitude DOUBLE NOT NULL,
  longitude DOUBLE NOT NULL,
  timezone VARCHAR NOT NULL,
  population BIGINT NOT NULL,
  elevation_m BIGINT,
  PRIMARY KEY (latitude, longitude)
);
```

```{sql connection=con_UFO}
COPY places FROM 'places.csv' (FORMAT CSV, HEADER, NULL 'NA');
```

The `day_parts_map` table contains NAs in `nautical_twilight_begin` and `astronomical_twilight_end`, so those columns allow nulls.

```{sql connection = con_UFO}
CREATE TABLE day_parts_map (
 rounded_lat INT NOT NULL,
 rounded_long INT NOT NULL,
 rounded_date DATE NOT NULL,
 astronomical_twilight_begin TIME,
 nautical_twilight_begin TIME,
 civil_twilight_begin TIME,
 sunrise TIME,
 solar_noon TIME,
 sunset TIME,
 civil_twilight_end TIME,
 nautical_twilight_end TIME,
 astronomical_twilight_end TIME,
  PRIMARY KEY (rounded_lat, rounded_long, rounded_date)
);
```

```{sql connection=con_UFO}
COPY day_parts_map FROM 'day_parts_map.csv' (FORMAT CSV, HEADER, NULL 'NA');
```

## Querying the data

### Query 1: Sightings by part of day

This query examines when sightings were reported most often by grouping reports by `day_part`, allowing us to compare counts across categories such as night, afternoon, morning, and the various twilight periods.

```{sql connection=con_UFO}
SELECT day_part, COUNT(*) AS sightings
FROM ufo_sightings
GROUP BY day_part
ORDER BY sightings DESC;
```

Sightings were most commonly reported at night, suggesting they were concentrated during darker conditions, when unusual lights or objects may be easier to notice. Time of day was clearly an important part of the reporting pattern in this dataset.

### Query 2: Population-adjusted sightings by state

This query compares sightings across U.S. states while adjusting for population. Rather than counting raw totals, it calculates sightings per million residents, giving a fairer comparison between large and small states.

```{sql connection=con_UFO}
WITH state_pop AS (
  SELECT state, SUM(population) AS pop
  FROM (
    SELECT DISTINCT city, state, country_code, population
    FROM places
    WHERE country_code = 'US'
      AND state IS NOT NULL
      AND population > 0
  )
  GROUP BY state
),
state_sightings AS (
  SELECT state, COUNT(*) AS sightings
  FROM ufo_sightings
  WHERE country_code = 'US'
    AND state IS NOT NULL
  GROUP BY state
)
SELECT s.state,
       s.sightings,
       p.pop,
       s.sightings * 1000000.0 / p.pop AS sightings_per_million
FROM state_sightings s
JOIN state_pop p
  ON s.state = p.state
ORDER BY sightings_per_million DESC;
```

Adjusting for population highlights states where sightings were unusually common relative to the number of residents. This is more informative than raw totals, since larger states naturally tend to accumulate more reports.

### Query 3: Reported shape by part of day

This query examines whether reported shapes varied by part of day. Grouping by both `day_part` and `shape` lets us compare how different types of sightings appeared across different lighting conditions.

```{sql connection=con_UFO}
SELECT day_part, shape, COUNT(*) AS count
FROM ufo_sightings
WHERE day_part IS NOT NULL
  AND shape IS NOT NULL
  AND shape IN ('light', 'circle', 'triangle', 'fireball', 'sphere', 'disk', 'oval', 'formation')
GROUP BY day_part, shape
ORDER BY count DESC;
```

The results connect what people reported seeing with when they reported it. If certain shapes appear more often at night or around dusk, that may suggest lighting conditions influenced how sightings were perceived and described.

### Query 4: Image documentation by shape

This query examines whether certain shapes were more likely to include images. For each shape it counts total sightings, counts how many included images, and calculates the percentage with images. We limit the results to shapes with at least 50 sightings so that rare categories do not distort the comparison.

```{sql connection=con_UFO}
SELECT shape,
  COUNT(*) AS total_sightings,
  SUM(CASE WHEN has_images THEN 1 ELSE 0 END) AS with_images,
  100.0 * SUM(CASE WHEN has_images THEN 1 ELSE 0 END) / COUNT(*) AS pct_with_images
FROM ufo_sightings
WHERE shape IS NOT NULL
GROUP BY shape
HAVING COUNT(*) >= 50
ORDER BY pct_with_images DESC;
```

This adds another layer to the analysis: it asks not only what people saw, but whether certain types of sightings were more likely to be supported by an image. Such differences may reflect variation in visibility, duration, or how distinct a given shape appeared to observers.

## Visualizations

### Sightings over time

This visualization examines how sightings changed over time by grouping reports by year, showing broader reporting trends across the dataset.

```{r}
sightings_by_year <- DBI::dbGetQuery(con_UFO, "
  SELECT EXTRACT(year FROM reported_date_time) AS year,
         COUNT(*) AS sightings
  FROM ufo_sightings
  WHERE reported_date_time IS NOT NULL
  GROUP BY year
  ORDER BY year;
")
ggplot(sightings_by_year, aes(x = year, y = sightings)) +
  geom_line() +
  geom_point() +
  labs(
    title = "UFO Sightings Over Time",
    x = "Year",
    y = "Number of Sightings"
  )
```

Sightings remained relatively low and stable through much of the early period, then increased sharply beginning in the late 1990s and early 2000s before peaking and declining in more recent years. This pattern suggests that reporting behavior or public interest changed considerably over time.

### Sightings by part of day (raw counts)

This visualization compares the number of reports across parts of the day, making it easier to see whether sightings were concentrated during darker periods or spread more evenly.

```{r}
daypart_counts <- dbGetQuery(con_UFO, "
  SELECT day_part, COUNT(*) AS sightings
  FROM ufo_sightings
  GROUP BY day_part
  ORDER BY sightings DESC;
")

ggplot(daypart_counts, aes(x = reorder(day_part, sightings), y = sightings)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "When UFO Sightings Happen Most Often",
    x = "Part of Day",
    y = "Number of Sightings"
  )
```

Sightings were overwhelmingly concentrated at night, with far more reports than any other period. The afternoon and dusk periods saw notable numbers as well, while morning and dawn were much lower. This again suggests sightings were more likely to be reported in darker conditions.

### Sightings by part of day (normalized per hour)

Raw counts are misleading because "night" spans far more hours than "dusk" or "dawn." To account for this, the query below normalizes sightings by the duration of each day-part at each rounded location and date, yielding sightings per available hour. The SQL for this normalization was drafted with the assistance of ChatGPT ([conversation](https://chatgpt.com/share/69ea9eed-59d4-83ea-98fd-9de727a5c9f1)), based on my approach of rounding the latitude and longitude in `places` to match `day_parts_map` and joining on rounded location and date.

```{r}
daypart_per_hour <- dbGetQuery(con_UFO, "
WITH sighting_rows AS (
  SELECT
      u.day_part,
      ROUND(p.latitude, -1)  AS rounded_lat,
      ROUND(p.longitude, -1) AS rounded_long,
      DATE(u.reported_date_time_utc) AS rounded_date
  FROM ufo_sightings u
  JOIN places p
    ON u.city = p.city
   AND u.state = p.state
   AND u.country_code = p.country_code
  WHERE u.day_part IS NOT NULL
),

counts AS (
  SELECT
      day_part,
      rounded_lat,
      rounded_long,
      rounded_date,
      COUNT(*) AS sightings
  FROM sighting_rows
  GROUP BY day_part, rounded_lat, rounded_long, rounded_date
),

durations AS (
  SELECT
      rounded_lat,
      rounded_long,
      rounded_date,

      (
        CASE
          WHEN EXTRACT(EPOCH FROM astronomical_twilight_end) < EXTRACT(EPOCH FROM astronomical_twilight_begin)
            THEN EXTRACT(EPOCH FROM astronomical_twilight_end) + 86400
          ELSE EXTRACT(EPOCH FROM astronomical_twilight_end)
        END
        - EXTRACT(EPOCH FROM astronomical_twilight_begin)
      ) / 3600.0 AS night_hours,

      (
        CASE
          WHEN EXTRACT(EPOCH FROM nautical_twilight_begin) < EXTRACT(EPOCH FROM astronomical_twilight_begin)
            THEN EXTRACT(EPOCH FROM nautical_twilight_begin) + 86400
          ELSE EXTRACT(EPOCH FROM nautical_twilight_begin)
        END
        - EXTRACT(EPOCH FROM astronomical_twilight_begin)
      ) / 3600.0 AS astronomical_dawn_hours,

      (
        CASE
          WHEN EXTRACT(EPOCH FROM civil_twilight_begin) < EXTRACT(EPOCH FROM nautical_twilight_begin)
            THEN EXTRACT(EPOCH FROM civil_twilight_begin) + 86400
          ELSE EXTRACT(EPOCH FROM civil_twilight_begin)
        END
        - EXTRACT(EPOCH FROM nautical_twilight_begin)
      ) / 3600.0 AS nautical_dawn_hours,

      (
        CASE
          WHEN EXTRACT(EPOCH FROM sunrise) < EXTRACT(EPOCH FROM civil_twilight_begin)
            THEN EXTRACT(EPOCH FROM sunrise) + 86400
          ELSE EXTRACT(EPOCH FROM sunrise)
        END
        - EXTRACT(EPOCH FROM civil_twilight_begin)
      ) / 3600.0 AS civil_dawn_hours,

      (
        CASE
          WHEN EXTRACT(EPOCH FROM solar_noon) < EXTRACT(EPOCH FROM sunrise)
            THEN EXTRACT(EPOCH FROM solar_noon) + 86400
          ELSE EXTRACT(EPOCH FROM solar_noon)
        END
        - EXTRACT(EPOCH FROM sunrise)
      ) / 3600.0 AS morning_hours,

      (
        CASE
          WHEN EXTRACT(EPOCH FROM sunset) < EXTRACT(EPOCH FROM solar_noon)
            THEN EXTRACT(EPOCH FROM sunset) + 86400
          ELSE EXTRACT(EPOCH FROM sunset)
        END
        - EXTRACT(EPOCH FROM solar_noon)
      ) / 3600.0 AS afternoon_hours,

      (
        CASE
          WHEN EXTRACT(EPOCH FROM civil_twilight_end) < EXTRACT(EPOCH FROM sunset)
            THEN EXTRACT(EPOCH FROM civil_twilight_end) + 86400
          ELSE EXTRACT(EPOCH FROM civil_twilight_end)
        END
        - EXTRACT(EPOCH FROM sunset)
      ) / 3600.0 AS civil_dusk_hours,

      (
        CASE
          WHEN EXTRACT(EPOCH FROM nautical_twilight_end) < EXTRACT(EPOCH FROM civil_twilight_end)
            THEN EXTRACT(EPOCH FROM nautical_twilight_end) + 86400
          ELSE EXTRACT(EPOCH FROM nautical_twilight_end)
        END
        - EXTRACT(EPOCH FROM civil_twilight_end)
      ) / 3600.0 AS nautical_dusk_hours,

      (
        CASE
          WHEN EXTRACT(EPOCH FROM astronomical_twilight_end) < EXTRACT(EPOCH FROM nautical_twilight_end)
            THEN EXTRACT(EPOCH FROM astronomical_twilight_end) + 86400
          ELSE EXTRACT(EPOCH FROM astronomical_twilight_end)
        END
        - EXTRACT(EPOCH FROM nautical_twilight_end)
      ) / 3600.0 AS astronomical_dusk_hours

  FROM day_parts_map
),

duration_long AS (
  SELECT rounded_lat, rounded_long, rounded_date, 'night' AS day_part, night_hours AS hours FROM durations
  UNION ALL
  SELECT rounded_lat, rounded_long, rounded_date, 'astronomical dawn', astronomical_dawn_hours FROM durations
  UNION ALL
  SELECT rounded_lat, rounded_long, rounded_date, 'nautical dawn', nautical_dawn_hours FROM durations
  UNION ALL
  SELECT rounded_lat, rounded_long, rounded_date, 'civil dawn', civil_dawn_hours FROM durations
  UNION ALL
  SELECT rounded_lat, rounded_long, rounded_date, 'morning', morning_hours FROM durations
  UNION ALL
  SELECT rounded_lat, rounded_long, rounded_date, 'afternoon', afternoon_hours FROM durations
  UNION ALL
  SELECT rounded_lat, rounded_long, rounded_date, 'civil dusk', civil_dusk_hours FROM durations
  UNION ALL
  SELECT rounded_lat, rounded_long, rounded_date, 'nautical dusk', nautical_dusk_hours FROM durations
  UNION ALL
  SELECT rounded_lat, rounded_long, rounded_date, 'astronomical dusk', astronomical_dusk_hours FROM durations
)

SELECT
    c.day_part,
    SUM(c.sightings) AS sightings,
    SUM(d.hours) AS total_available_hours,
    SUM(c.sightings) / SUM(d.hours) AS sightings_per_hour
FROM counts c
JOIN duration_long d
  ON c.rounded_lat = d.rounded_lat
 AND c.rounded_long = d.rounded_long
 AND c.rounded_date = d.rounded_date
 AND c.day_part = d.day_part
GROUP BY c.day_part
ORDER BY sightings_per_hour DESC;
")
```

```{r}
ggplot(daypart_per_hour, aes(x = reorder(day_part, sightings), y = sightings_per_hour)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "When UFO Sightings Happen Most Often",
    x = "Part of Day",
    y = "Number of Sightings Per Hour"
  )
```

After normalizing for the share of the day each part occupies, dusk and dawn show a much higher density of sightings than night, morning, or afternoon. Dusk and dawn are also the times of poorest visibility: low sun angles create glare, and rapidly changing light levels make it harder for the eye to adjust, reducing peripheral vision and depth perception. This may help explain the elevated number of sightings reported during these periods.

```{r}
#| include: false
# Close the connection and shut down DuckDB so the process releases cleanly.
DBI::dbDisconnect(con_UFO, shutdown = TRUE)
```
