---
title: "SQL project"
---

**NOTE: This is joint work with Joseph Dee.**
github can be found at https://github.com/ADC-405-S26/Project_2_SQL_Team_13/tree/main

# load packages and connect to database server

```{r setup}
#| message: false
#| warning: false
#| include: false
library(tidyverse)
library(dbplyr)
library(RMariaDB)
library(duckdb) # R package for building a duckdb database locally.

#  put comment on the code/codechunks that we wrote

# install and load more packages if required
# install.packages("package_name")
# library(package_name)
```
# Aiden

### ---------------------------------------------------------------------------
# Accessing data
Data was taken from here https://github.com/rfordatascience/tidytuesday/blob/main/data/2023/2023-06-20/readme.md
There are three .csv files "ufo_sightings.csv" "places.csv" "day_parts_map.csv"
ufo_sightings.csv has city and state in common with places.csv
ufo_sightings.csv has reported_date_utc and posted data in common with day_parts_map.csv

- Aiden

```{r}

raw_ufo_sightings <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2023/2023-06-20/ufo_sightings.csv')
raw_places <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2023/2023-06-20/places.csv')
day_parts_map <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2023/2023-06-20/day_parts_map.csv')

glimpse(raw_ufo_sightings) #shape has lots of NAs and day_part has some NAs
glimpse(raw_places) #alternative_city_names & elevation_m has NAs
glimpse(day_parts_map) # No NAs

#places <- raw_places %>% slice(-167,-986,-3904,-989,-815,-992,-8302,-69,-1315) # Daytona Beach (and some others) appear multiple times
places <- raw_places %>% distinct(latitude, longitude, .keep_all = TRUE) # there are ~625 duplicate longitude AND latitude entries
ufo_sightings <- raw_ufo_sightings %>% distinct(reported_date_time_utc, city, .keep_all = TRUE) #we want to use reported_date_time_utc as key

write_csv(ufo_sightings, "ufo_sightings.csv")  
write_csv(places, "places.csv")
write_csv(day_parts_map, "day_parts_map.csv") 
```
# Aiden

### ---------------------------------------------------------------------------
# Designing the database

```{r}
con_UFO <- DBI::dbConnect(duckdb::duckdb(), dbdir = "UFO_db")
```
# Aiden

```{sql connection = con_UFO}
DROP TABLE IF EXISTS ufo_sightings;
DROP TABLE IF EXISTS places;
DROP TABLE IF EXISTS day_parts_map;
```
To get the right times refer here https://duckdb.org/docs/current/sql/data_types/overview
column | type | nulls allowed? | default value
I am not sure what we should set the default values too so I have not included them at this stage

Currently there is no primary key for this table. I have emailed professor abhishek about what to do.
# Aiden
```{sql connection = con_UFO}
CREATE TABLE ufo_sightings (
  reported_date_time TIMESTAMP NOT NULL,
  reported_date_time_utc TIMESTAMP NOT NULL,
  posted_date DATE NOT NULL,
  city VARCHAR NOT NULL,
  state VARCHAR,
  country_code VARCHAR NOT NULL,
  shape VARCHAR,
  reported_duration VARCHAR NOT NULL,
  duration_seconds BIGINT NOT NULL,
  summary VARCHAR,
  has_images BOOLEAN NOT NULL,
  day_part VARCHAR,
  PRIMARY KEY (reported_date_time_utc, city) 
);
```
# Aiden

```{sql connection=con_UFO}

COPY ufo_sightings FROM 'ufo_sightings.csv' (FORMAT CSV, HEADER, NULL 'NA');

```
# Aiden

```{sql connection = con_UFO}
CREATE TABLE places (
  city VARCHAR NOT NULL,
  alternate_city_names VARCHAR,
  state VARCHAR, 
  country VARCHAR NOT NULL,
  country_code VARCHAR NOT NULL,
  latitude DOUBLE NOT NULL,
  longitude DOUBLE NOT NULL,
  timezone VARCHAR NOT NULL,
  population BIGINT NOT NULL,
  elevation_m BIGINT,
  PRIMARY KEY (latitude, longitude) 
);
```
# Aiden

```{sql connection=con_UFO}

COPY places FROM 'places.csv' (FORMAT CSV, HEADER, NULL 'NA');

```
# Aiden
day_parts_map has NAs within "nautical_twilight_begin" and "astronomical_twilight_end"
```{sql connection = con_UFO}
CREATE TABLE day_parts_map (
 rounded_lat INT NOT NULL,
 rounded_long INT NOT NULL,
 rounded_date DATE NOT NULL,
 astronomical_twilight_begin TIME,
 nautical_twilight_begin TIME,
 civil_twilight_begin TIME,
 sunrise TIME,
 solar_noon TIME,
 sunset TIME,
 civil_twilight_end TIME,
 nautical_twilight_end TIME,
 astronomical_twilight_end TIME,
  PRIMARY KEY (rounded_lat, rounded_long, rounded_date) 
);
```
# Aiden

```{sql connection=con_UFO}

COPY day_parts_map FROM 'day_parts_map.csv' (FORMAT CSV, HEADER, NULL 'NA');

```
# Aiden
### ---------------------------------------------------------------------------
# Querying data using SQL

1. This query examines when UFO sightings happened most often by grouping reports by day_part. This allowed us to compare the number of sightings across categories such as night, afternoon, morning, and different twilight periods.

```{sql connection=con_UFO}

SELECT day_part, COUNT(*) AS sightings
FROM ufo_sightings
GROUP BY day_part
ORDER BY sightings DESC;

```
# Jeb

The results show that UFO sightings were most commonly reported at night. This suggests that sightings were concentrated during darker conditions, when unusual lights or objects may have been easier to notice. It also shows that time of day was an important part of the reporting pattern in this dataset.





2. This query compares UFO sightings across U.S. states while adjusting for population. Instead of only counting total sightings, it calculates sightings per million residents, which gives a fairer comparison between large and small states.
```{sql connection=con_UFO}
WITH state_pop AS (
  SELECT state, SUM(population) AS pop
  FROM (
    SELECT DISTINCT city, state, country_code, population
    FROM places
    WHERE country_code = 'US'
      AND state IS NOT NULL
      AND population > 0
  )
  GROUP BY state
),
state_sightings AS (
  SELECT state, COUNT(*) AS sightings
  FROM ufo_sightings
  WHERE country_code = 'US'
    AND state IS NOT NULL
  GROUP BY state
)
SELECT s.state,
       s.sightings,
       p.pop,
       s.sightings * 1000000.0 / p.pop AS sightings_per_million
FROM state_sightings s
JOIN state_pop p
  ON s.state = p.state
ORDER BY sightings_per_million DESC;


```
The population-adjusted results help identify states where sightings were unusually common relative to the number of people living there. This is more useful than raw totals because larger states naturally tend to have more reports. By adjusting for population, the query highlights states with higher relative reporting rates.


# Jeb

3. This query examines whether reported UFO shapes varied by part of day. By grouping sightings by both day_part and shape, we could compare how different types of sightings appeared across different lighting conditions.
```{sql connection=con_UFO}

SELECT day_part, shape, COUNT(*) AS count
FROM ufo_sightings
WHERE day_part IS NOT NULL
  AND shape IS NOT NULL
  AND shape IN ('light', 'circle', 'triangle', 'fireball', 'sphere', 'disk', 'oval', 'formation')
GROUP BY day_part, shape
ORDER BY count DESC;
```
The results show how reported shapes differed across parts of the day. This helps connect what people reported seeing with when they reported seeing it. If certain shapes appeared more often at night or around dusk, that may suggest that lighting conditions affected how sightings were perceived and described.


4. This query examines whether certain UFO shapes were more likely to include images. For each reported shape, it counts the total number of sightings, counts how many included images, and calculates the percentage of sightings with images. We limited the results to shapes with at least 50 sightings so that very rare categories would not distort the comparison.
```{sql connection=con_UFO}

SELECT shape,
  COUNT(*) AS total_sightings,
  SUM(CASE WHEN has_images THEN 1 ELSE 0 END) AS with_images,
  100.0 * SUM(CASE WHEN has_images THEN 1 ELSE 0 END) / COUNT(*) AS pct_with_images
FROM ufo_sightings
WHERE shape IS NOT NULL
GROUP BY shape
HAVING COUNT(*) >= 50
ORDER BY pct_with_images DESC;
```
The results help show whether some UFO shapes were more likely to be visually documented than others. This adds another layer to the analysis because it does not just ask what people saw, but whether certain types of sightings were more likely to be supported by images. This may reflect differences in visibility, duration, or how distinct certain shapes appeared to observers.

2. This query compares UFO sightings across U.S. states while adjusting for population. Instead of only counting total sightings, it calculates sightings per million residents, which gives a fairer comparison between large and small states.
```{sql connection=con_UFO}
WITH state_pop AS (
  SELECT state, SUM(population) AS pop
  FROM (
    SELECT DISTINCT city, state, country_code, population
    FROM places
    WHERE country_code = 'US'
      AND state IS NOT NULL
      AND population > 0
  )
  GROUP BY state
),
state_sightings AS (
  SELECT state, COUNT(*) AS sightings
  FROM ufo_sightings
  WHERE country_code = 'US'
    AND state IS NOT NULL
  GROUP BY state
)
SELECT s.state,
       s.sightings,
       p.pop,
       s.sightings * 1000000.0 / p.pop AS sightings_per_million
FROM state_sightings s
JOIN state_pop p
  ON s.state = p.state
ORDER BY sightings_per_million DESC;


```
The population-adjusted results help identify states where sightings were unusually common relative to the number of people living there. This is more useful than raw totals because larger states naturally tend to have more reports. By adjusting for population, the query highlights states with higher relative reporting rates.


# Jeb

3. This query examines whether reported UFO shapes varied by part of day. By grouping sightings by both day_part and shape, we could compare how different types of sightings appeared across different lighting conditions.
```{sql connection=con_UFO}

SELECT day_part, shape, COUNT(*) AS count
FROM ufo_sightings
WHERE day_part IS NOT NULL
  AND shape IS NOT NULL
  AND shape IN ('light', 'circle', 'triangle', 'fireball', 'sphere', 'disk', 'oval', 'formation')
GROUP BY day_part, shape
ORDER BY count DESC;
```
The results show how reported shapes differed across parts of the day. This helps connect what people reported seeing with when they reported seeing it. If certain shapes appeared more often at night or around dusk, that may suggest that lighting conditions affected how sightings were perceived and described.


4. This query examines whether certain UFO shapes were more likely to include images. For each reported shape, it counts the total number of sightings, counts how many included images, and calculates the percentage of sightings with images. We limited the results to shapes with at least 50 sightings so that very rare categories would not distort the comparison.
```{sql connection=con_UFO}

SELECT shape,
  COUNT(*) AS total_sightings,
  SUM(CASE WHEN has_images THEN 1 ELSE 0 END) AS with_images,
  100.0 * SUM(CASE WHEN has_images THEN 1 ELSE 0 END) / COUNT(*) AS pct_with_images
FROM ufo_sightings
WHERE shape IS NOT NULL
GROUP BY shape
HAVING COUNT(*) >= 50
ORDER BY pct_with_images DESC;
```
The results help show whether some UFO shapes were more likely to be visually documented than others. This adds another layer to the analysis because it does not just ask what people saw, but whether certain types of sightings were more likely to be supported by images. This may reflect differences in visibility, duration, or how distinct certain shapes appeared to observers.

### ---------------------------------------------------------------------------
# Creating data visualizations

This visualization examines how UFO sightings changed over time by grouping reports by year. The graph helps show broader reporting trends across the dataset rather than focusing only on one category, location, or time of day.
```{r}
sightings_by_year <- DBI::dbGetQuery(con_UFO, "
  SELECT EXTRACT(year FROM reported_date_time) AS year,
         COUNT(*) AS sightings
  FROM ufo_sightings
  WHERE reported_date_time IS NOT NULL
  GROUP BY year
  ORDER BY year;
")
ggplot(sightings_by_year, aes(x = year, y = sightings)) +
  geom_line() +
  geom_point() +
  labs(
    title = "UFO Sightings Over Time",
    x = "Year",
    y = "Number of Sightings"
  ) 
```
This visualization shows how the number of reported UFO sightings changed over time. Sightings remained relatively low and stable for much of the early period, but began to increase sharply starting in the late 1990s and early 2000s. The number of sightings then reached a peak before declining in more recent years. This pattern suggests that reporting behavior or public interest may have changed significantly over time.





This visualization shows when UFO sightings were reported most often by comparing the number of reports across parts of the day. The graph makes it easier to see whether sightings were concentrated during darker periods or spread more evenly across the day.
```{r}
daypart_counts <- dbGetQuery(con_UFO, "
  SELECT day_part, COUNT(*) AS sightings
  FROM ufo_sightings
  GROUP BY day_part
  ORDER BY sightings DESC;
")

ggplot(daypart_counts, aes(x = reorder(day_part, sightings), y = sightings)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "When UFO Sightings Happen Most Often",
    x = "Part of Day",
    y = "Number of Sightings"
  )

```

This visualization shows when UFO sightings were reported most often by comparing the number of reports across different parts of the day. Sightings were overwhelmingly concentrated at night, with significantly more reports than any other time period. There were also notable numbers of sightings during the afternoon and dusk periods, while morning and dawn categories were much lower. This pattern suggests that sightings were more likely to be reported in darker conditions, when unusual lights or objects may be easier to observe.


- Aiden Below
To more accurately display during which parts of the day UFO sightings are commononly spotted we need to normalize the data according to the portion of time "night" "sunset" takes up, we see that night has a ton of obversations, its also much longer than "sunset". If we round the lat and long values within "places" to match the lat long values within "day_parts_map" we can then get the duration of each event at each location on the date specified within "ufo_sightings" under "reported_date_time_utc" as well as what day_parts_map category it fell under "night" "dusk" etc. I knew it was possible to do this in SQL but I could not get it to work so I explained my thoughts to chatgpt so it would write the actual SQL code for my reasoning. Code below written by ChatGPT (https://chatgpt.com/share/69ea9eed-59d4-83ea-98fd-9de727a5c9f1)

```{r}
daypart_per_hour <- dbGetQuery(con_UFO, "
WITH sighting_rows AS (
  SELECT
      u.day_part,
      ROUND(p.latitude, -1)  AS rounded_lat,
      ROUND(p.longitude, -1) AS rounded_long,
      DATE(u.reported_date_time_utc) AS rounded_date
  FROM ufo_sightings u
  JOIN places p
    ON u.city = p.city
   AND u.state = p.state
   AND u.country_code = p.country_code
  WHERE u.day_part IS NOT NULL
),

counts AS (
  SELECT
      day_part,
      rounded_lat,
      rounded_long,
      rounded_date,
      COUNT(*) AS sightings
  FROM sighting_rows
  GROUP BY day_part, rounded_lat, rounded_long, rounded_date
),

durations AS (
  SELECT
      rounded_lat,
      rounded_long,
      rounded_date,

      (
        CASE
          WHEN EXTRACT(EPOCH FROM astronomical_twilight_end) < EXTRACT(EPOCH FROM astronomical_twilight_begin)
            THEN EXTRACT(EPOCH FROM astronomical_twilight_end) + 86400
          ELSE EXTRACT(EPOCH FROM astronomical_twilight_end)
        END
        - EXTRACT(EPOCH FROM astronomical_twilight_begin)
      ) / 3600.0 AS night_hours,

      (
        CASE
          WHEN EXTRACT(EPOCH FROM nautical_twilight_begin) < EXTRACT(EPOCH FROM astronomical_twilight_begin)
            THEN EXTRACT(EPOCH FROM nautical_twilight_begin) + 86400
          ELSE EXTRACT(EPOCH FROM nautical_twilight_begin)
        END
        - EXTRACT(EPOCH FROM astronomical_twilight_begin)
      ) / 3600.0 AS astronomical_dawn_hours,

      (
        CASE
          WHEN EXTRACT(EPOCH FROM civil_twilight_begin) < EXTRACT(EPOCH FROM nautical_twilight_begin)
            THEN EXTRACT(EPOCH FROM civil_twilight_begin) + 86400
          ELSE EXTRACT(EPOCH FROM civil_twilight_begin)
        END
        - EXTRACT(EPOCH FROM nautical_twilight_begin)
      ) / 3600.0 AS nautical_dawn_hours,

      (
        CASE
          WHEN EXTRACT(EPOCH FROM sunrise) < EXTRACT(EPOCH FROM civil_twilight_begin)
            THEN EXTRACT(EPOCH FROM sunrise) + 86400
          ELSE EXTRACT(EPOCH FROM sunrise)
        END
        - EXTRACT(EPOCH FROM civil_twilight_begin)
      ) / 3600.0 AS civil_dawn_hours,

      (
        CASE
          WHEN EXTRACT(EPOCH FROM solar_noon) < EXTRACT(EPOCH FROM sunrise)
            THEN EXTRACT(EPOCH FROM solar_noon) + 86400
          ELSE EXTRACT(EPOCH FROM solar_noon)
        END
        - EXTRACT(EPOCH FROM sunrise)
      ) / 3600.0 AS morning_hours,

      (
        CASE
          WHEN EXTRACT(EPOCH FROM sunset) < EXTRACT(EPOCH FROM solar_noon)
            THEN EXTRACT(EPOCH FROM sunset) + 86400
          ELSE EXTRACT(EPOCH FROM sunset)
        END
        - EXTRACT(EPOCH FROM solar_noon)
      ) / 3600.0 AS afternoon_hours,

      (
        CASE
          WHEN EXTRACT(EPOCH FROM civil_twilight_end) < EXTRACT(EPOCH FROM sunset)
            THEN EXTRACT(EPOCH FROM civil_twilight_end) + 86400
          ELSE EXTRACT(EPOCH FROM civil_twilight_end)
        END
        - EXTRACT(EPOCH FROM sunset)
      ) / 3600.0 AS civil_dusk_hours,

      (
        CASE
          WHEN EXTRACT(EPOCH FROM nautical_twilight_end) < EXTRACT(EPOCH FROM civil_twilight_end)
            THEN EXTRACT(EPOCH FROM nautical_twilight_end) + 86400
          ELSE EXTRACT(EPOCH FROM nautical_twilight_end)
        END
        - EXTRACT(EPOCH FROM civil_twilight_end)
      ) / 3600.0 AS nautical_dusk_hours,

      (
        CASE
          WHEN EXTRACT(EPOCH FROM astronomical_twilight_end) < EXTRACT(EPOCH FROM nautical_twilight_end)
            THEN EXTRACT(EPOCH FROM astronomical_twilight_end) + 86400
          ELSE EXTRACT(EPOCH FROM astronomical_twilight_end)
        END
        - EXTRACT(EPOCH FROM nautical_twilight_end)
      ) / 3600.0 AS astronomical_dusk_hours

  FROM day_parts_map
),

duration_long AS (
  SELECT rounded_lat, rounded_long, rounded_date, 'night' AS day_part, night_hours AS hours FROM durations
  UNION ALL
  SELECT rounded_lat, rounded_long, rounded_date, 'astronomical dawn', astronomical_dawn_hours FROM durations
  UNION ALL
  SELECT rounded_lat, rounded_long, rounded_date, 'nautical dawn', nautical_dawn_hours FROM durations
  UNION ALL
  SELECT rounded_lat, rounded_long, rounded_date, 'civil dawn', civil_dawn_hours FROM durations
  UNION ALL
  SELECT rounded_lat, rounded_long, rounded_date, 'morning', morning_hours FROM durations
  UNION ALL
  SELECT rounded_lat, rounded_long, rounded_date, 'afternoon', afternoon_hours FROM durations
  UNION ALL
  SELECT rounded_lat, rounded_long, rounded_date, 'civil dusk', civil_dusk_hours FROM durations
  UNION ALL
  SELECT rounded_lat, rounded_long, rounded_date, 'nautical dusk', nautical_dusk_hours FROM durations
  UNION ALL
  SELECT rounded_lat, rounded_long, rounded_date, 'astronomical dusk', astronomical_dusk_hours FROM durations
)

SELECT
    c.day_part,
    SUM(c.sightings) AS sightings,
    SUM(d.hours) AS total_available_hours,
    SUM(c.sightings) / SUM(d.hours) AS sightings_per_hour
FROM counts c
JOIN duration_long d
  ON c.rounded_lat = d.rounded_lat
 AND c.rounded_long = d.rounded_long
 AND c.rounded_date = d.rounded_date
 AND c.day_part = d.day_part
GROUP BY c.day_part
ORDER BY sightings_per_hour DESC;
")
```


```{r}
ggplot(daypart_per_hour, aes(x = reorder(day_part, sightings), y = sightings_per_hour)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "When UFO Sightings Happen Most Often",
    x = "Part of Day",
    y = "Number of Sightings Per Hour"
  )
```

After normalizing for the portion of day that each part of the day occupies we can see that there is a much higher density of detections during dusk and dawn compared to night morning and night. Dusk and Dawn are the worst times for visibility. During dawn and dusk low sun angles create blinding glare, while rapidly changing light levels make it difficult for the human eye to adjust, reducing peripheral vision and depth perception. This likely explains the large number of UFO sightings seen during these times.


