---
title: "Text Analysis Project: UFO Sighting Summaries"
toc: true
---

*Joint work with Joseph Dee. Source code is available on [GitHub](https://github.com/ADC-405-S26/Project_3_TextAnalysis_Team_13/tree/main).*

## Introduction

The data comes from the National UFO Reporting Center (NUFORC), which holds more than 95,000 unidentified flying object (UFO) reports from around the world. We enriched it with data from sunrise-sunset.org to label the part of the day (dawn, morning, night, and so on) for each sighting's date, latitude, and longitude. NUFORC reports can be submitted anonymously online; the earliest online reports date to 1998, though historical sightings reach back to 1925. This version of the data was scraped in June 2023 and can be found [here](https://github.com/rfordatascience/tidytuesday/blob/main/data/2023/2023-06-20/readme.md). Before submitting, users are shown a list of common misidentifications (Starlink satellite launches, Jupiter, and the like), and reports are "validated" before being added to the database, although the rigor of these checks is unclear.

We began by tokenizing the summaries to find the most frequently occurring words across the entire dataset. The 20 most common words appear in the table below.

```{r}
#| code-fold: true
#| echo: false
#| message: false
#| warning: false
library(tidyverse)
library(scales)
library(tidytext)
library(textdata)
library(mdsr)
library(knitr)

# Import data
ufo_sightings <- read_csv("ufo_sightings.csv", show_col_types = FALSE)

# Count the most common words and format a labeled table
most_common <- ufo_sightings %>%
    unnest_tokens(word, summary) %>%
    anti_join(stop_words, by = "word") %>%
    count(word, sort = TRUE) %>%
    mutate(percent = n / sum(n)) %>%
    slice_head(n = 20) %>%
    rename(Word = word,
          Occurrences = n,
          Percentage = percent) %>%
    mutate(Occurrences = comma(Occurrences),
          Percentage = percent(Percentage, accuracy = 0.01))

kable(most_common,
      caption = "Most Common Words in UFO Sighting Summaries",
      align = c("l", "r", "r"))
```

Unsurprisingly, the most common word was "lights," followed closely by "light" and "sky." More interestingly, three colors appear in the top ten: "orange," "white," and "red." This made me wonder whether different colors might be more common during different parts of the day—I would have guessed, for instance, that "white" would be more common at night than in the afternoon.

## Color mentions by time of day

```{r}
#| code-fold: true
# Extract color mentions (lowercased first, since some appear in all caps),
# drop rows with no color or no day_part, then plot color share by day part.
color_data <- ufo_sightings %>%
    mutate(color = str_extract(str_to_lower(summary),"red|green|blue|orange|yellow|silver|white|black|purple|pink|gold|metallic")) %>% 
    filter(!is.na(color), !is.na(day_part)) %>% count(day_part, color)

ggplot(color_data, aes(x = day_part, y = n, fill = color)) +
    geom_col(position = "fill") +
    scale_fill_manual(values = c(red = "red",green = "green",blue = "blue",orange = "orange", yellow = "yellow", silver = "gray70", white = "white", black = "black",purple = "purple",pink = "hotpink",gold = "gold",metallic = "#999B9B")) +
    scale_y_continuous(labels = percent) +
    labs(title = "Color Mentions by Time of Day", x = "Time of Day", y = "Proportion", fill = "Color") +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

My expectation was wrong: "white" is actually mentioned more during the morning and afternoon than at night. The table below shows the underlying numbers for each color across times of day.

```{r}
#| code-fold: true
# Table formatting generated with assistance from ChatGPT.
library(DT)

color_table <- color_data %>%
    group_by(day_part) %>%
    mutate(
        proportion = n / sum(n)
    ) %>%
    ungroup() %>%
    arrange(day_part, desc(proportion)) %>%
    rename(
        day_part = day_part,
        Color = color,
        Mentions = n,
        Proportion = proportion
    ) %>%
    mutate(
        Mentions = comma(Mentions),
        Proportion = percent(Proportion, accuracy = 0.1)
    )

datatable(
    color_table,
    rownames = FALSE,
    options = list(
        pageLength = 12,
        autoWidth = TRUE
    ),
    caption = "Color Mentions by Time of Day"
)
```

## Most common bigrams

Next I looked at bigrams—pairs of consecutive words—to find the most common two-word phrases.

```{r}
#| code-fold: true
# Remove stopwords from the bigrams (adapted with assistance from ChatGPT),
# then drop automated phrases that the site appends rather than user text.
ufo_bigrams_nostopwords <- ufo_sightings %>%
    unnest_tokens(
        output = bigram,
        input = summary,
        token = "ngrams",
        n = 2
    ) %>%
    separate(
        bigram,
        into = c("word1", "word2"),
        sep = " "
    ) %>%
    filter(
        !word1 %in% tidytext::stop_words$word,
        !word2 %in% tidytext::stop_words$word
    ) %>%
    unite(
        bigram,
        word1,
        word2,
        sep = " "
    ) %>%
    filter(
        !bigram %in% c(
            "nuforc note",
            "anonymous report"
        )
    )

bigram_table <- ufo_bigrams_nostopwords %>%
    count(bigram, sort = TRUE) %>%
    slice_head(n = 25) %>%
    rename(
        Bigram = bigram,
        Occurrences = n
    ) %>%
    mutate(
        Occurrences = comma(Occurrences)
    )

kable(
    bigram_table,
    caption = "Most Common Bigrams in UFO Sighting Summaries",
    align = c("l", "r")
)
```

Two of the most common individual words form the most common bigram, "bright light." More interesting are the 1,090 occurrences of "starlink satellites." Searching by single word would only have surfaced "satellites," which tells a different story than the bigram does. The site warns users that what they saw may have been a Starlink satellite launch, and rather than omitting these entries it appends "(starlink satellites?)" to the submitted summary when it suspects this is the cause. Having watched a Starlink launch cross the night sky myself, I was completely convinced I had seen several UFOs until I remembered what such launches look like: a string of perhaps 40 bright white lights moving in unison across the sky.

## Distinctive words by time of day

I was curious which words are most distinctive to each time of day. The table below shows the ten most distinctive words for each period. A high TF-IDF score marks a word that is unusually specific to one time of day and not common across the others. Some standouts include "fireworks" at nautical dusk, but on the whole I found the single-word results underwhelming.

```{r}
#| code-fold: true
tfidf_table <- ufo_sightings %>%
    filter(!is.na(summary), !is.na(day_part)) %>%
    unnest_tokens(word, summary) %>%
    anti_join(stop_words, by = "word") %>%
    count(day_part, word, sort = TRUE) %>%
    bind_tf_idf(word, day_part, n) %>%
    group_by(day_part) %>%
    slice_max(tf_idf, n = 10, with_ties = FALSE) %>%
    ungroup() %>%
    arrange(day_part, desc(tf_idf)) %>%
    rename(
        `Time of Day` = day_part,
        Word = word,
        Count = n,
        `TF-IDF` = tf_idf
    ) %>%
    mutate(
        Count = comma(Count),
        `TF-IDF` = round(`TF-IDF`, 4)
    )

kable(
    tfidf_table,
    caption = "Most Distinctive UFO Summary Words by Time of Day",
    align = c("l", "l", "r", "r")
)
```

## Distinctive bigrams by time of day

Applying the same approach to bigrams yielded much more interesting results. At nautical dusk there are many military-related phrases such as "missile launch" and "navy missile," along with "fireworks display." Most striking to me is that "silver sphere" is strongly associated with the morning and afternoon but largely absent at other times of day.

```{r}
#| code-fold: true
bigram_tfidf_table <- ufo_bigrams_nostopwords %>%
    ungroup() %>%
    filter(!is.na(day_part), !is.na(bigram)) %>%
    count(day_part, bigram, name = "n") %>%
    bind_tf_idf(bigram, day_part, n) %>%
    group_by(day_part) %>%
    slice_max(tf_idf, n = 10, with_ties = FALSE) %>%
    ungroup() %>%
    arrange(day_part, desc(tf_idf)) %>%
    rename(
        `Time of Day` = day_part,
        Bigram = bigram,
        Count = n,
        `TF-IDF` = tf_idf
    ) %>%
    mutate(
        Count = comma(Count),
        `TF-IDF` = round(`TF-IDF`, 4)
    )

kable(
    bigram_tfidf_table,
    caption = "Most Distinctive UFO Summary Bigrams by Time of Day",
    align = c("l", "l", "r", "r")
)
```

## Sentiment analysis

To better understand the tone of the reports, we conducted sentiment analysis on the witness summaries. After tokenizing the text, we matched terms against the Bing sentiment lexicon to classify them as positive or negative, then assigned each report a sentiment score based on the balance of positive and negative words in its summary. We compared these scores across sighting-duration groups and across times of day to see whether descriptions varied with the time of a sighting or with how long an encounter lasted.

### Sentiment by part of day

```{r}
#| code-fold: true
# Sentiment analysis adapted with assistance from ChatGPT.
ufo_sentiment <- ufo_sightings %>%
    filter(!is.na(summary), !is.na(day_part)) %>%
    
    unnest_tokens(word, summary) %>%
    
    inner_join(
        get_sentiments("bing"),
        by = "word"
        )

ufo_sentiment_counts <- ufo_sentiment %>%
    count(day_part, sentiment)

ggplot(ufo_sentiment_counts, aes(x = day_part, y = n, fill = sentiment)) +
    geom_col(position = "fill") +
    scale_y_continuous(labels = scales::percent) +
    labs(title = "Sentiment by Day Part",
         x = "Day Part",
         y = "Proportion",
         fill = "Sentiment") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

This visualization compares the proportion of positive and negative language across parts of the day. Sentiment stayed relatively balanced throughout, with some noticeable differences: reports during astronomical dawn, astronomical dusk, civil dawn, and nautical dawn contained a slightly higher proportion of positive language, while morning and afternoon reports skewed somewhat more negative. One explanation is that sightings in darker or transitional light may have inspired more curiosity and wonder, whereas daytime sightings—seen in full daylight and harder to explain away—may have felt more credible and unsettling, prompting more negative language. Cold War–era anxieties about military technology, surveillance, and unidentified aircraft may have heightened that concern. Although the differences are modest, the analysis suggests that environmental conditions and visibility influence not only what witnesses believe they saw, but also how they emotionally describe it.

### Sentiment over time

```{r}
ufo_sentiment_year <- ufo_sightings %>%
    filter(!is.na(summary), !is.na(reported_date_time)) %>%
    mutate(
        year = lubridate::year(reported_date_time),
        report_id = row_number()
    ) %>%
    filter(year >= 1950, year <= 2022) %>%
    unnest_tokens(word, summary) %>%
    inner_join(get_sentiments("bing"), by = "word") %>%
    mutate(sentiment_value = if_else(sentiment == "positive", 1, -1)) %>%
    group_by(report_id, year) %>%
    summarize(sentiment_score = sum(sentiment_value), .groups = "drop") %>%
    group_by(year) %>%
    summarize(
        avg_sentiment = mean(sentiment_score),
        num_reports = n(),
        .groups = "drop"
    ) %>%
    filter(num_reports >= 20)

ggplot(ufo_sentiment_year,
       aes(x = year, y = avg_sentiment)) +
    geom_line() +
    geom_point(aes(size = num_reports), alpha = 0.6) +
    geom_hline(yintercept = 0, linetype = "dashed") +
    scale_size_continuous(name = "Number of Reports") +
    labs(
        title = "How Has UFO Sentiment Changed Over Time?",
        subtitle = "Average sentiment score of UFO sighting summaries by year",
        x = "Year",
        y = "Average Sentiment Score"
    ) +
    theme_minimal()

```

The average sentiment of UFO reports gradually became more positive over time, particularly after the late 1990s. Earlier reports tended toward more neutral or slightly negative language, while more recent sightings were increasingly described with words associated with curiosity, amazement, and fascination. One explanation is a shift in public perception: during the Cold War, UFO sightings were often tied to fears about military technology, surveillance, and national security, whereas modern discussion of UFOs is more commonly associated with entertainment and curiosity, which may encourage less fearful, more intrigued descriptions.

The sharp increase in reports after the late 1990s should be interpreted carefully, however, because of a key limitation in the data. In 1998, NUFORC opened its online reporting system to the public, allowing individuals to submit their own sightings directly. As a result, the data before 1998 consists of historical reports archived after the fact rather than submitted in real time, so the earlier years reflect a smaller and more selective record, while the later years capture a much larger volume of firsthand submissions. Even with this limitation, the visualization still suggests a broader shift in how people emotionally frame UFO encounters over time.

### Sentiment by sighting duration

```{r}
ufo_report_sentiment <- ufo_sightings %>%
    filter(!is.na(summary), !is.na(duration_seconds)) %>%
    mutate(
        report_id = row_number(),
        duration_group = case_when(
            duration_seconds < 60 ~ "Under 1 minute",
            duration_seconds < 300 ~ "1 to 5 minutes",
            duration_seconds < 900 ~ "5 to 15 minutes",
            duration_seconds >= 900 ~ "15+ minutes"
        )
    ) %>%
    unnest_tokens(word, summary) %>%
    inner_join(get_sentiments("bing"), by = "word") %>%
    mutate(sentiment_value = if_else(sentiment == "positive", 1, -1)) %>%
    group_by(report_id, duration_group) %>%
    summarize(sentiment_score = sum(sentiment_value), .groups = "drop")

ufo_sentiment_bars <- ufo_report_sentiment %>%
    count(duration_group, sentiment_score) %>%
    group_by(duration_group) %>%
    mutate(prop = n / sum(n)) %>%
    ungroup()

ggplot(ufo_sentiment_bars,
       aes(x = factor(sentiment_score),
           y = prop,
           fill = duration_group)) +
    geom_col(show.legend = FALSE) +
    facet_wrap(~ duration_group, ncol = 1) +
    scale_y_continuous(labels = scales::percent) +
    labs(
        title = "Do Longer UFO Sightings Sound More Negative?",
        subtitle = "Sentiment scores from tokenized witness summaries",
        x = "Sentiment Score",
        y = "Share of Reports"
    ) +
    theme_minimal()
```

Across all duration categories, the sentiment distributions remained remarkably similar, with most reports clustered around neutral to slightly positive scores regardless of how long the sighting lasted. Even sightings of more than fifteen minutes did not produce noticeably more negative language than those under a minute. This suggests that witnesses generally described encounters in a calm, observational manner rather than with fear or panic—contrary to the popular-culture portrayal of prolonged encounters as emotionally intense. One possibility is that many witnesses approached sightings with curiosity rather than alarm; another is that people writing for a database like NUFORC adopted a deliberately factual tone to appear credible, focusing on documenting what they observed as clearly and objectively as possible.
