---
title: "text analysis project"
---



**NOTE: This is joint work with Joseph Dee**
github can be found at https://github.com/ADC-405-S26/Project_3_TextAnalysis_Team_13/tree/main

# Introduction to our data.

Data was taken from The National UFO Reporting Center (NUFORC) which has \>95,000 unidentified flying object (UFO) reports from all over the globe. Our data was enriched with sunrise-sunset.org data to give information about what part of the day (dawn, morning, night, etc) it was on that day, in that general latitude and longitude. UFO reports can be anonymous and submitted online with the earliest reports posted in 1998, however, historical sightings are also included back to 1925. This data was scraped in June of 2023 and found [here](https://github.com/rfordatascience/tidytuesday/blob/main/data/2023/2023-06-20/readme.md). Before submitting a report users are given a list of common misidentifications (starlink satellite launches, Jupiter, etc) and reports are "validated" before being added to the database though I am not sure of the rigor of these checks.

First we tokenized all the words to extract the most commonly occuring words within the entire dataset. The top 1000 words can be found in the table below.

```{r}
#| code-fold: true
#| echo: false
library(tidyverse)
library(dplyr)
library(scales)
library(tidytext)
library(textdata)
library(mdsr)
library(knitr)
library(scales)

# Data is imported
ufo_sightings <- read_csv("ufo_sightings.csv",show_col_types = FALSE)

# Most common words
most_common <- ufo_sightings %>%
    unnest_tokens(word, summary) %>%
    anti_join(tidytext::get_stopwords(), by = "word") %>%
    count(word, sort = TRUE)
# Make the table with good labels
most_common <- ufo_sightings %>%
    unnest_tokens(word, summary) %>%
    anti_join(stop_words, by = "word") %>%
    count(word, sort = TRUE) %>%
    mutate(percent = n / sum(n)) %>%
    slice_head(n = 20) %>%
    rename(Word = word,
          Occurrences = n,
          Percentage = percent) %>%
    mutate(Occurrences = comma(Occurrences),
          Percentage = percent(Percentage, accuracy = 0.01))

kable(most_common,
      caption = "Most Common Words in UFO Sighting Summaries",
      align = c("l", "r", "r"))
```

Understandable the most common word was "lights", followed closely by "light" and "sky". Interestingly we see three colors in the ten most common "orange", "white" and "red". This made me wonder if these different colors could be more common during different parts of the day. I would imagine that "white" is more common in the nighttime compared to the afternoon.

```{r}
#| code-fold: true
# extract colors (some were in all caps so make them lowercase first)
# then make a graph that displays those specific colors organized by daypart
# also remove all NA's from colors and NA's from day_part
color_data <- ufo_sightings %>%
    mutate(color = str_extract(str_to_lower(summary),"red|green|blue|orange|yellow|silver|white|black|purple|pink|gold|metallic")) %>% 
    filter(!is.na(color), !is.na(day_part)) %>% count(day_part, color)

ggplot(color_data, aes(x = day_part, y = n, fill = color)) +
    geom_col(position = "fill") +
    scale_fill_manual(values = c(red = "red",green = "green",blue = "blue",orange = "orange", yellow = "yellow", silver = "gray70", white = "white", black = "black",purple = "purple",pink = "hotpink",gold = "gold",metallic = "#999B9B")) +
    scale_y_continuous(labels = percent) +
    labs(title = "Color Mentions by Time of Day", x = "Time of Day", y = "Proportion", fill = "Color") +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

Interestingly I was proven wrong, it appears white is actually mentioned more during the morning and afternoon than it is at night. I have also included a table with the numerical data for each of these colors during different times of the day below.

```{r}
#| code-fold: true
# table code generated by chatGPT
library(DT)

color_table <- color_data %>%
    group_by(day_part) %>%
    mutate(
        proportion = n / sum(n)
    ) %>%
    ungroup() %>%
    arrange(day_part, desc(proportion)) %>%
    rename(
        day_part = day_part,
        Color = color,
        Mentions = n,
        Proportion = proportion
    ) %>%
    mutate(
        Mentions = comma(Mentions),
        Proportion = percent(Proportion, accuracy = 0.1)
    )

datatable(
    color_table,
    rownames = FALSE,
    options = list(
        pageLength = 12,
        autoWidth = TRUE
    ),
    caption = "Color Mentions by Time of Day"
)
```

Next I wanted to look at bigrams, or groups of two words and what would be the most common group of two words.

```{r}
#| code-fold: true
# Wanted to remove stopwords from bigram GPT wrote this below, after being given my code without stopwords removed
ufo_bigrams_nostopwords <- ufo_sightings %>%
    unnest_tokens(
        output = bigram,
        input = summary,
        token = "ngrams",
        n = 2
    ) %>%
    separate(
        bigram,
        into = c("word1", "word2"),
        sep = " "
    ) %>%
    filter(
        !word1 %in% tidytext::stop_words$word,
        !word2 %in% tidytext::stop_words$word
    ) %>%
    unite(
        bigram,
        word1,
        word2,
        sep = " "
    ) %>%
    filter(
        !bigram %in% c(
            "nuforc note",
            "anonymous report"
        )
    )

bigram_table <- ufo_bigrams_nostopwords %>%
    count(bigram, sort = TRUE) %>%
    slice_head(n = 25) %>%
    rename(
        Bigram = bigram,
        Occurrences = n
    ) %>%
    mutate(
        Occurrences = comma(Occurrences)
    )

kable(
    bigram_table,
    caption = "Most Common Bigrams in UFO Sighting Summaries",
    align = c("l", "r")
)
# things like "nuforc note" and "anonymous report" are not interesting as they are automated inclusions by the site
```

Unsurprisingly two of the most common words make up the most common bigram "bright light". However, more interesting is the 1,090 inclusions of "starlink satellites" searching up by just word would only have given us "satellites" which gives a different story than the bigram. Interestingly the site warns users that what they are seeing could easily be a starlink satellite launch. Instead of ommitting these entries the site seems to append ((starlink satellites?)) to the end of the user submitted summary if they suspect this is what the reporter saw. Having seen a starlink satellite launch travelling through the sky at night I 100% was convinced I had seen several UFOs until remembering that is just what starlink satellite launches look like. A string of maybe 40 bright white lights moving in unison across the night sky.

I was curious what the most unqiue words are during different times of day. The table below shows each of the times of days as well as their 10 most unique phrase. A row with a high TF-IDF represents a single word that is unique to that time of day and not commonly found in other times of days. Some standouts are "fireworks" appearing frequently at nautical dusk, but I was largely unhappy with the results when looking at single words.

```{r}
#| code-fold: true
tfidf_table <- ufo_sightings %>%
    filter(!is.na(summary), !is.na(day_part)) %>%
    unnest_tokens(word, summary) %>%
    anti_join(stop_words, by = "word") %>%
    count(day_part, word, sort = TRUE) %>%
    bind_tf_idf(word, day_part, n) %>%
    group_by(day_part) %>%
    slice_max(tf_idf, n = 10, with_ties = FALSE) %>%
    ungroup() %>%
    arrange(day_part, desc(tf_idf)) %>%
    rename(
        `Time of Day` = day_part,
        Word = word,
        Count = n,
        `TF-IDF` = tf_idf
    ) %>%
    mutate(
        Count = comma(Count),
        `TF-IDF` = round(`TF-IDF`, 4)
    )

kable(
    tfidf_table,
    caption = "Most Distinctive UFO Summary Words by Time of Day",
    align = c("l", "l", "r", "r")
)
```

Next I looked at two word groups to find which was the most common at different times of day. Which yielded much more interesting results, at nautical dusk there are lots of military related phrases like "missile launch" "navy missle", as well as "fireworks display". Most interesting to me is the fact that "silver sphere" is highly correlated with the morning and afternoon but not seen during other times of day.

```{r}
#| code-fold: true
bigram_tfidf_table <- ufo_bigrams_nostopwords %>%
    ungroup() %>%
    filter(!is.na(day_part), !is.na(bigram)) %>%
    count(day_part, bigram, name = "n") %>%
    bind_tf_idf(bigram, day_part, n) %>%
    group_by(day_part) %>%
    slice_max(tf_idf, n = 10, with_ties = FALSE) %>%
    ungroup() %>%
    arrange(day_part, desc(tf_idf)) %>%
    rename(
        `Time of Day` = day_part,
        Bigram = bigram,
        Count = n,
        `TF-IDF` = tf_idf
    ) %>%
    mutate(
        Count = comma(Count),
        `TF-IDF` = round(`TF-IDF`, 4)
    )

kable(
    bigram_tfidf_table,
    caption = "Most Distinctive UFO Summary Bigrams by Time of Day",
    align = c("l", "l", "r", "r")
)
```
To better understand the emotions and tones of the UFO reports, we conducted sentiment analysis on the witness testimonies and summaries provided in the data set. We took the tokenized words, we matched those terms to the Bing sentiment lexicon in order to classify the terms into positive and negative tones, in hopes to understand the sentiment of the observers. Each report was then assigned a sentiment score based on the balance of positive and negative words that appear in the sighting summary. We compared these sentiment scores across different sighting duration groups as well as different times of the day. We did this to determine whether people have positive or negative reviews depending on the time of day, and to determine whether longer encounters were described differently than shorter ones.

```{r}
#| code-fold: true
# sentiment analysis 
# GPT assisted due to missing class going over trump tweets


ufo_sentiment <- ufo_sightings %>%
    filter(!is.na(summary), !is.na(day_part)) %>%
    
    unnest_tokens(word, summary) %>%
    
    inner_join(
        get_sentiments("bing"),
        by = "word"
        )

ufo_sentiment_counts <- ufo_sentiment %>%
    count(day_part, sentiment)

ggplot(ufo_sentiment_counts, aes(x = day_part, y = n, fill = sentiment)) +
    geom_col(position = "fill") +
    scale_y_continuous(labels = scales::percent) +
    labs(title = "Sentiment by Day Part",
         x = "Day Part",
         y = "Proportion",
         fill = "Sentiment") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
This specific visualization compares the proportion of positive and negative language used in UFO sighting summaries across different parts of the day. The sentiment of UFO reports remained relatively balanced across all time periods, although there were some noticeable differences. Reports occurring during astronomical dawn, astronomical dusk, civil dawn, and nautical dawn contained a slightly higher proportion of positive language. The reports from the morning and afternoon showed a greater share of negative language. One possible explanation is that sightings occurring in darker or transitional lighting conditions may have inspired more curiosity and wonder, while sightings during brighter daylight hours may have been described with more confusion and concern. One possible theory is that daytime UFO sightings caused more concern because objects seen in full daylight were often viewed as more credible and harder to explain away. During the Cold War, fears surrounding military technology, surveillance, and unidentified aircraft may have also increased public anxiety about unusual objects in the sky. As a result, witnesses describing daytime sightings may have used more negative language because the encounters felt more real or unsettling. Although the differences were not extreme, the visualization demonstrates that the language used in UFO reports changes subtly depending on the time of day. This suggests that environmental conditions and visibility may influence not only what witnesses believe they saw, but also how they emotionally describe the experience.

```{r}
ufo_sentiment_year <- ufo_sightings %>%
    filter(!is.na(summary), !is.na(reported_date_time)) %>%
    mutate(
        year = lubridate::year(reported_date_time),
        report_id = row_number()
    ) %>%
    filter(year >= 1950, year <= 2022) %>%
    unnest_tokens(word, summary) %>%
    inner_join(get_sentiments("bing"), by = "word") %>%
    mutate(sentiment_value = if_else(sentiment == "positive", 1, -1)) %>%
    group_by(report_id, year) %>%
    summarize(sentiment_score = sum(sentiment_value), .groups = "drop") %>%
    group_by(year) %>%
    summarize(
        avg_sentiment = mean(sentiment_score),
        num_reports = n(),
        .groups = "drop"
    ) %>%
    filter(num_reports >= 20)

ggplot(ufo_sentiment_year,
       aes(x = year, y = avg_sentiment)) +
    geom_line() +
    geom_point(aes(size = num_reports), alpha = 0.6) +
    geom_hline(yintercept = 0, linetype = "dashed") +
    scale_size_continuous(name = "Number of Reports") +
    labs(
        title = "How Has UFO Sentiment Changed Over Time?",
        subtitle = "Average sentiment score of UFO sighting summaries by year",
        x = "Year",
        y = "Average Sentiment Score"
    ) +
    theme_minimal()

```
As stated above, the average sentiment of UFO reports gradually became more positive over time, particularly after the late 1990s. Earlier reports generally contained more neutral or slightly negative language, while more recent sightings were increasingly described with words associated with curiosity, amazement, and fascination. One possible theory is that public perceptions of UFOs changed over time. During the Cold War era, UFO sightings were often connected to fears surrounding military technology, surveillance, and national security. In contrast, modern discussions of UFOs are more commonly tied to entertainment and curiosity. This may encourage witnesses to describe sightings in a less fearful and more intrigued manner.

At the same time, the sharp increase in reports after the late 1990s should be interpreted carefully because of an important limitation in the dataset. In 1998, the National UFO Reporting Center opened its online reporting system to the public, allowing individuals to directly submit their own sightings. As a result, the data before 1998 consists of historical reports that were archived after the fact rather than submitted in real time. This means the earlier years reflect a smaller and more selective historical record. At the same time the later years capture a much larger volume of firsthand public submissions. Even with this limitation, the visualization still suggests a broader shift in how people emotionally frame UFO encounters over time.

```{r}
ufo_report_sentiment <- ufo_sightings %>%
    filter(!is.na(summary), !is.na(duration_seconds)) %>%
    mutate(
        report_id = row_number(),
        duration_group = case_when(
            duration_seconds < 60 ~ "Under 1 minute",
            duration_seconds < 300 ~ "1 to 5 minutes",
            duration_seconds < 900 ~ "5 to 15 minutes",
            duration_seconds >= 900 ~ "15+ minutes"
        )
    ) %>%
    unnest_tokens(word, summary) %>%
    inner_join(get_sentiments("bing"), by = "word") %>%
    mutate(sentiment_value = if_else(sentiment == "positive", 1, -1)) %>%
    group_by(report_id, duration_group) %>%
    summarize(sentiment_score = sum(sentiment_value), .groups = "drop")

ufo_sentiment_bars <- ufo_report_sentiment %>%
    count(duration_group, sentiment_score) %>%
    group_by(duration_group) %>%
    mutate(prop = n / sum(n)) %>%
    ungroup()

ggplot(ufo_sentiment_bars,
       aes(x = factor(sentiment_score),
           y = prop,
           fill = duration_group)) +
    geom_col(show.legend = FALSE) +
    facet_wrap(~ duration_group, ncol = 1) +
    scale_y_continuous(labels = scales::percent) +
    labs(
        title = "Do Longer UFO Sightings Sound More Negative?",
        subtitle = "Sentiment scores from tokenized witness summaries",
        x = "Sentiment Score",
        y = "Share of Reports"
    ) +
    theme_minimal()
```

Across all duration categories, the sentiment distributions remained remarkably similar. Most UFO reports clustering around neutral to slightly positive scores regardless of how long the sighting lasted. Even sightings lasting more than fifteen minutes did not produce dramatically more negative language than sightings lasting under a minute. This pattern suggests that witnesses generally described UFO encounters in a calm and observational manner rather than with extreme fear or panic. This finding goes against many conventional beliefs surrounding UFO sightings. Popular culture often portrays prolonged encounters as emotionally intense experiences associated with fear. One possible theory is that many witnesses approached sightings with curiosity rather than alarm. Another possibility is that witnesses writing reports for databases such as NUFORC intentionally adopted a more factual tone in order to appear credible. Instead of writing emotionally charged narratives, many individuals may have focused on documenting what they believed they observed as clearly and objectively as possible.

